---
title: "Gene Expression"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
We aim to find the transcriptional signatures associated with a particular condition.

## Objective
In this Rmarkdown document, we will explore the results of the Differential Expression Analysis (DEA) between the different levels associated to a particular condition.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(rstatix)
library(UCell)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_r_objects_in <- here::here("03_clustering_annotation/results/R_objects")
path_r_results <- here::here("04_analysis/3_geneExpression/results")

# Functions
source(here::here("bin/utils.R"))

cells_color_palette <- c(
                     "Basal keratinocytes"= "#E6B800",
                     "Supra-Basal keratinocytes"= "#F4E409",
                     "Granular keratinocytes"= "#FF6000", 
                     "Cornified keratinocytes"= "#FFA500",
                     "Dead keratinocytes"= "#583101",
                      "Cycling basal keratinocytes"= "#FF944D")

cells_list <- c( "Basal keratinocytes",
           "Supra-Basal keratinocytes",
           "Granular keratinocytes",
           "Cornified keratinocytes",
           "Dead keratinocytes",
           "Cycling basal keratinocytes"
                                       )

# Pseudobulk function (equivalent to DecoupleR)
get_pseudobulk <- function(seurat_obj, sample_col = "individual", groups_col = "cell_type", min_cells = 0, min_counts = 0) {
  # Extract raw integer counts
  counts_matrix <- GetAssayData(seurat_obj, slot = "counts")
  # Extract metadata
  metadata <- seurat_obj@meta.data
  
  # Create a combined group identifier
  metadata$group_id <- paste(metadata[[sample_col]], metadata[[groups_col]], sep = "_")
  
  # Group cells by the combined identifier
  cell_groups <- split(rownames(metadata), metadata$group_id)
  # Aggregate counts by summing across all cells in each group
  pseudobulk_counts <- sapply(cell_groups, function(cells) {
    if (length(cells) >= min_cells) {
      summed_counts <- Matrix::rowSums(counts_matrix[, cells, drop = FALSE])
      if (sum(summed_counts) >= min_counts) {
        return(summed_counts)
      }
    }
    return(rep(0, nrow(counts_matrix)))
  })
  # Convert to matrix format
  pseudobulk_counts <- as.matrix(pseudobulk_counts)
  # Remove empty columns
  valid_cols <- colSums(pseudobulk_counts) > 0
  pseudobulk_counts <- pseudobulk_counts[, valid_cols, drop = FALSE]
  
  # Create sample-level metadata
  sample_metadata <- unique(metadata[, c(sample_col, groups_col), drop = FALSE])
  sample_metadata$group_id <- paste(sample_metadata[[sample_col]], sample_metadata[[groups_col]], sep = "_")
  rownames(sample_metadata) <- sample_metadata$group_id
  # Filter metadata to match valid columns
  sample_metadata <- sample_metadata[valid_cols, , drop = FALSE]
  
  # Create a new Seurat object with pseudobulk data
  pseudobulk_seurat <- CreateSeuratObject(
    counts = pseudobulk_counts,
    meta.data = sample_metadata
  )
  
  return(pseudobulk_seurat)
}
```


```{r}
## Load data
seurat_obj <- readRDS(paste0(path_r_objects_in, "/post12_cd45neg_alldata_cleaned_final_clustering_annotation_Zeb1.rds"))
seurat_obj
```

```{r}
seurat_obj_sub <- seurat_obj[, (seurat_obj$celltypes_1 %in% c("Epidermal cells"))]

seurat_obj_sub$celltypes_2 <- as.character(seurat_obj_sub$celltypes_2)
seurat_obj_sub$celltypes_2 <- factor(x = seurat_obj_sub$celltypes_2,
                                     levels = cells_list)
seurat_obj_sub
table(seurat_obj_sub$sample_id, seurat_obj_sub$celltypes_2)

Idents(seurat_obj_sub) <- "celltypes_2"
```

# SPRR gene family
```{r}
SPRR_genes <- sort(rownames(seurat_obj_sub)[grep("^Sprr", rownames(seurat_obj_sub))])
SPRR_genes
```

## Across populations
```{r fig.width=16, fig.height=8}
VlnPlot(seurat_obj_sub,
        features = SPRR_genes,
        group.by = "celltypes_2",
        cols =  cells_color_palette,
        ncol = 7)
```


```{r}
seurat_obj_sub$celltypes_2 <- factor(x = seurat_obj_sub$celltypes_2,
                                     levels =rev(cells_list))
Idents(seurat_obj_sub) <- "celltypes_2"

dotplot_cells <- DotPlot(seurat_obj_sub,
        features = SPRR_genes,
        cols = "RdBu",
        scale = FALSE,
        dot.scale = 5) +
  #coord_flip() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5, face = "italic"))

dotplot_cells

pdf(file = paste0(path_r_results, "/cd45neg_Zeb1_Epidermis_dotplot_SPRRfamily.pdf"))
print(dotplot_cells)    
dev.off()   
```

## Across condition

```{r fig.width=16, fig.height=8}
VlnPlot(seurat_obj_sub,
        features = SPRR_genes,
        group.by = "sample_id",
        cols =  c("#FEFAE0", "#DDA15E", "#BC6C25"),
        ncol = 7)
```

### Statistics
```{r}
# Store all stats
all_stats <- list()

for (gene in SPRR_genes) {
  
  # Create dataframe
  vln_df <- data.frame(
    expression = seurat_obj_sub[["RNA"]]@data[gene, ], 
    condition = seurat_obj_sub$sample_id
  )
  
  # Set condition factor levels
  vln_df$condition <- factor(
    x = vln_df$condition,
    levels = c("Z1_WT_CD45neg", "Z1_WT_IMQ_CD45neg", "Z1_KO_IMQ_CD45neg")
  )
  
  vln_df$expression <- as.numeric(vln_df$expression)
  
  # Define comparisons
  mycomparisons <- list(
    c("Z1_WT_IMQ_CD45neg", "Z1_WT_CD45neg"),
    c("Z1_KO_IMQ_CD45neg", "Z1_WT_IMQ_CD45neg")
  )

  # Optional count summary
  vln_df_count <- vln_df %>%
    group_by(condition) %>%
    summarise(diff_0 = sum(expression != 0), .groups = "drop")
  print(paste("Counts for gene:", gene))

  # Wilcoxon test across conditions
  stat.test <- wilcox_test(
    vln_df,
    expression ~ condition,
    comparisons = mycomparisons,
    alternative = "two.sided"
  ) %>%
    adjust_pvalue(method = "fdr") %>%
    add_significance()
  
  stat.test <- stat.test %>% add_xy_position()
  stat.test$gene <- gene  # tag with gene name

  all_stats[[gene]] <- stat.test  # store result
}

# Combine all results and save
combined_stats <- bind_rows(all_stats)
combined_stats$groups <- NULL

write.table(combined_stats,
            file = "results/cd45neg_Zeb1_Epidermis_statistics_SPRRfamily_Condition.csv",
            sep = ",", row.names = FALSE)

# View all results
combined_stats
```


## Across population & condition

### Signature

The function AddModuleScore_UCell() allows operating directly on Seurat objects. UCell scores are calculated from raw counts or normalized data, and returned as metadata columns. The example below defines some simple signatures, and applies them on single-cell data stored in a Seurat object.

#### Pseudobulk
```{r}
# Pseudobulk
seurat_obj_Epi_pseudo <- get_pseudobulk(seurat_obj_sub, sample_col = "sample_id", groups_col = "celltypes_2", min_cells = 0, min_counts = 0)
seurat_obj_Epi_pseudo

## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj_Epi_pseudo <- AddModuleScore_UCell(seurat_obj_Epi_pseudo, 
                                   features=list("SPRR_genes" = SPRR_genes), name=NULL)
head(seurat_obj_Epi_pseudo@meta.data)
```

```{r fig.height=4, fig.width=5}
gene_signatures_scores_df <- seurat_obj_Epi_pseudo@meta.data[, c("celltypes_2", "sample_id", "SPRR_genes")]

gene_signatures_scores_df$celltypes_2 <- factor(gene_signatures_scores_df$celltypes_2, levels = cells_list)
gene_signatures_scores_df <- gene_signatures_scores_df %>%
  arrange(celltypes_2, sample_id)

heatmap_data <- gene_signatures_scores_df %>%
  pivot_wider(names_from = sample_id, values_from = SPRR_genes) %>%
  column_to_rownames("celltypes_2")

write_csv(gene_signatures_scores_df,
          file = "results/cd45neg_Zeb1_Epidermis_Heatmap_pseudoSPRRsignature.csv")

pheatmap_cell_by_cond <- pheatmap::pheatmap(t(heatmap_data), 
         scale = "none",
         cluster_rows = FALSE, # Cluster rows
         cluster_cols = FALSE, # Cluster columns
         show_rownames = TRUE, # Show row names
         show_colnames = TRUE, # Show column names
         #breaks = seq(0, 1, length.out = 101), # Legend from 0 to 1
         #color = colorRampPalette(c("blue", "white", "red"))(100), # Color scale
         border_color = "white",
         main = "SPRR Gene Signatures")


pdf("results/cd45neg_Zeb1_Epidermis_Heatmap_pseudoSPRRsignature.pdf", width = 5, height = 4, onefile = TRUE)
print(pheatmap_cell_by_cond)
dev.off()
```

#### Cell level & statistics
```{r}
## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj_sub <- AddModuleScore_UCell(seurat_obj_sub, 
                                   features=list("SPRR_genes" = SPRR_genes), name=NULL)
head(seurat_obj_sub@meta.data)
```

```{r fig.height=3.5, fig.width=5}
gene_signatures_scores <- seurat_obj_sub@meta.data[, c("celltypes_2", "sample_id", "SPRR_genes")]

gene_signatures_scores_df <- gene_signatures_scores %>%
  group_by(sample_id, celltypes_2) %>%
  summarise(mean_SPRR = mean(SPRR_genes, na.rm = TRUE), .groups = "drop")
head(gene_signatures_scores_df)

gene_signatures_scores_df$celltypes_2 <- factor(gene_signatures_scores_df$celltypes_2, levels = cells_list)
gene_signatures_scores_df <- gene_signatures_scores_df %>%
  arrange(celltypes_2, sample_id)

heatmap_data <- gene_signatures_scores_df %>%
  pivot_wider(names_from = sample_id, values_from = mean_SPRR) %>%
  column_to_rownames("celltypes_2")

write_csv(gene_signatures_scores_df,
          file = "results/cd45neg_Zeb1_Epidermis_Heatmap_SPRRsignature.csv")

pheatmap_cell_by_cond <- pheatmap::pheatmap(t(heatmap_data), 
         scale = "none",
         cluster_rows = FALSE, # Cluster rows
         cluster_cols = FALSE, # Cluster columns
         show_rownames = TRUE, # Show row names
         show_colnames = TRUE, # Show column names
         #breaks = seq(0, 1, length.out = 101), # Legend from 0 to 1
         #color = colorRampPalette(c("blue", "white", "red"))(100), # Color scale
         border_color = "white",
         main = "SPRR Gene Signatures")


pdf("results/cd45neg_Zeb1_Epidermis_Heatmap_SPRRsignature.pdf", width = 5, height = 3.5, onefile = TRUE)
print(pheatmap_cell_by_cond)
dev.off()
```

```{r}
mycomparisons <- list(c("Z1_WT_IMQ_CD45neg", "Z1_WT_CD45neg"), c("Z1_KO_IMQ_CD45neg", "Z1_WT_IMQ_CD45neg"))

# Statistics for signature
  vln_df <- data.frame(
    SPRR_genes = seurat_obj_sub$SPRR_genes, 
    cells = seurat_obj_sub$celltypes_2,
    condition = seurat_obj_sub$sample_id
  )
vln_df$SPRR_genes <- as.numeric(vln_df$SPRR_genes)

# Factor levels for condition
vln_df$cells <- factor(vln_df$cells, 
                           levels = cells_list)
vln_df$condition <- factor(vln_df$condition, 
                           levels = c("Z1_WT_CD45neg", "Z1_WT_IMQ_CD45neg", "Z1_KO_IMQ_CD45neg"))

stat.test <- vln_df %>%
  group_by(cells) %>%
  wilcox_test(SPRR_genes ~ condition, comparisons = mycomparisons, alternative = "two.sided") %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance()

write.table(stat.test, 
            file = "results/cd45neg_Zeb1_Epidermis_Heatmap_SPRRsignature_statistics.csv", 
            sep = ",", row.names = FALSE)

# Optional: View combined results
stat.test
```


```{r fig.width=17, fig.height=20}
SPRR_genes_subset <- c("Sprr1a", "Sprr1b", "Sprr2b" , "Sprr2d", "Sprr2e", "Sprr2f", "Sprr2h", "Sprr2i","Sprr2j-ps")

vln_df <- seurat_obj_sub[["RNA"]]@data[SPRR_genes_subset, ] %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  tidyr::pivot_longer(
    cols = -gene,
    names_to = "cell_id",
    values_to = "expression"
  )

# Add metadata (cell types and condition)
vln_df$cells <- seurat_obj_sub$celltypes_2[match(vln_df$cell_id, colnames(seurat_obj_sub))]
vln_df$condition <- seurat_obj_sub$sample_id[match(vln_df$cell_id, colnames(seurat_obj_sub))]


# Factor levels for condition
vln_df$cells <- factor(vln_df$cells, 
                           levels = cells_list)
vln_df$condition <- factor(vln_df$condition, 
                           levels = c("Z1_WT_CD45neg", "Z1_WT_IMQ_CD45neg", "Z1_KO_IMQ_CD45neg"))

# Plot
gg_vln <- vln_df %>% 
  ggplot(aes(x = condition, y = expression, fill = condition)) + 
  geom_violin(adjust =1, trim=TRUE, scale = "width") +
  facet_wrap(cells ~ gene, scales = "free_y", ncol = 9) +
  scale_fill_manual(values = c("#FEFAE0", "#DDA15E", "#BC6C25")) +
  geom_jitter(alpha=0.4) +
  labs(x = "", y = "Expression Levels", fill = "Condition") +
  theme_minimal() +
  theme(strip.text = element_text(size = 15),
        axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.text = element_text(size=14),
        legend.position = "bottom") +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15)))
print(gg_vln)

pdf("results/cd45neg_Zeb1_Epidermis_VlnPlot_SPRRfamily.pdf", width = 17, height = 20, onefile = TRUE)
print(gg_vln)
dev.off()
```

### Statistics

```{r}
# Initialize an empty list to collect all stats
all_stats <- list()

for (gene in SPRR_genes_subset) {
  
  # Create violin plot dataframe for the gene
  vln_df <- data.frame(
    expression = seurat_obj_sub[["RNA"]]@data[gene, ], 
    cells = seurat_obj_sub$celltypes_2,
    condition = seurat_obj_sub$sample_id
  )
  
  vln_df$condition <- factor(
    x = vln_df$condition,
    levels = c("Z1_WT_CD45neg", "Z1_WT_IMQ_CD45neg", "Z1_KO_IMQ_CD45neg")
  )
  
  vln_df$expression <- as.numeric(vln_df$expression)

  mycomparisons <- list(
    c("Z1_WT_IMQ_CD45neg", "Z1_WT_CD45neg"), 
    c("Z1_KO_IMQ_CD45neg", "Z1_WT_IMQ_CD45neg")
  )

  # Check counts (optional for debug)
  vln_df_count <- vln_df %>% 
    group_by(cells, condition) %>% 
    summarise(diff_0 = sum(expression != 0), .groups = "drop")
  print(paste("Counts for gene:", gene))

  # Compute statistics
  stat.test <- vln_df %>%
    group_by(cells) %>%
    wilcox_test(expression ~ condition, comparisons = mycomparisons, alternative = "two.sided") %>%
    adjust_pvalue(method = "fdr") %>%
    add_significance()
  
  stat.test <- stat.test %>% add_xy_position()
  stat.test$gene <- gene  # Add gene info

  all_stats[[gene]] <- stat.test  # Save per gene stats
  
  # Optional: save each gene's result individually
  # write.table(as.data.frame(stat.test),
  #             file = paste0("results/statistics_", gene, ".csv"),
  #             sep = ",", row.names = FALSE)
}

# Combine and save all stats
combined_stats <- bind_rows(all_stats)
combined_stats$groups <- NULL

write.table(combined_stats, 
            file = "results/cd45neg_Zeb1_Epidermis_statistics_SPRRfamily_CellsCondition.csv", 
            sep = ",", row.names = FALSE)

# Optional: View combined results
combined_stats
```


# Session info
```{r}
sessionInfo()
```
