---
title: "Annotation to subcluster: CD45pos"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
After the pre-processing steps done in previous notebooks, we will now start with the downstream data analysis in order to extract biological insights and describe the underlying biology.

## Objective
In this Rmarkdown document, we are going to assign cell type clusters into different cell-types clusters based on the similarity of their gene expression profiles. Then, we will perform again process each object to define cell-type specific clusters with more resolution.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(DT)
library(dittoSeq)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/POST_12_metadata.csv")
path_samples_metadata <- here::here("01_cellranger_mapping/data/POST_12_metadata_demultiplexing.csv")
path_r_objects <- here::here("03_clustering_annotation/results/R_objects")
path_r_tables <- here::here("03_clustering_annotation/results/tables")
path_r_figs <- here::here("03_clustering_annotation/results/figs")

# Functions
source(here::here("bin/utils.R"))

## Define resolution of choice
resolution <- 0.1
resolution_name <- paste0("RNA_snn_res.", resolution)

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(35, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **POST_12** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata_samples <- read.csv(path_samples_metadata)

# Merging metadata_all tables
metadata <- metadata %>% filter(type == "cDNA" & fraction == "CD45-")
metadata_all <- left_join(metadata, subset(metadata_samples, select = -c(gem_id)), by = "library_name")

print("Libraries/Samples metadata")
DT::datatable(metadata_all, options = list(scrollX = TRUE))

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/post12_cd45pos_clustering_resolutions.rds"))
DefaultAssay(seurat_obj) <- "RNA"
Idents(seurat_obj) <- "RNA_snn_res.0.1"
seurat_obj

# Load biomarkers
seurat_obj_markers <- readRDS(paste0(path_r_objects, "/biomarkers_cd45pos_resolution-0.1.rds"))

# Modify data
seurat_obj$sample_id <- factor(seurat_obj$sample_id,
                               levels = c("Z1_WT_CD45pos",
                                          "Z2_WT_CD45pos",
                                          "Z1_WT_IMQ_CD45pos",
                                          "Z2_WT_IMQ_CD45pos",
                                          "Z1_KO_IMQ_CD45pos",
                                          "Z2_KO_IMQ_CD45pos"
                                          ))
```

# Explore cell clusters
To visualize the cell clusters, there are a few different dimensionality reduction techniques, such as t-SNE and UMAP. These methods require to input a number of PCA dimensions (same number as the one used for the clustering). We will not run UMAP again as it was computed in the previous notebook.
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering <- DimPlot(object = seurat_obj,
        group.by = resolution_name,
        label = TRUE,
        label.size = 6,
        cols = color_palette,
        pt.size=0.1
        ) & NoLegend()
gg_umap_clustering
```

Now, we will also explore clusters across conditions.
```{r fig.width=20, fig.height=5}
gg_clustering_by_cond <- DimPlot(
  object = seurat_obj,
  group.by = resolution_name,
  split.by = "sample_id",
  label = FALSE,
  cols = color_palette
  )

gg_clustering_by_cond
```


## Cell clusters data exploration
```{r}
# Number of cells / cluster
table(seurat_obj$RNA_snn_res.0.1)

# Number of cells per cluster vs confounder variables
table(seurat_obj$sample_id, seurat_obj$RNA_snn_res.0.1)
```


## Assessing sources of uninteresting variation
Next, we will explore additional metrics, such as the number of UMIs and genes per cell, mitochondrial / ribosomal gene expression and scrublet doublet scores on the UMAP embedding. 

```{r fig.width=14, fig.height=10}
qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb")

gg_qc_umap <- FeaturePlot(
  seurat_obj,
  reduction = "umap",
  features = qc_metrics,
  cols = c("lightgrey", "darkblue"), 
  order = TRUE,
  ncol = 2)

gg_qc_umap
```

```{r fig.width=14, fig.height=15}
gg_qc_vln <- VlnPlot(
  seurat_obj,
  features = qc_metrics,
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_qc_vln
```

Finally, we will also check the cell-cycle phase by assessing S-phase and G2M-phase markers signature.
```{r fig.width=14, fig.height=10}
gg_phase_vln <- VlnPlot(
  seurat_obj,
  features = c("S.Score", "G2M.Score"),
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_phase_vln
```

# Explore cluster biomarkers
```{r}
DefaultAssay(seurat_obj) <- "RNA"
```

## Table
```{r}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

top_all_markers
DT::datatable(top_all_markers, options = list(scrollX = TRUE))
```

## Dotplot
```{r fig.width=20, fig.height=5}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC)

top_genemarkers_list <- unique(top_all_markers$gene)

DotPlot(
  seurat_obj,
  features = top_genemarkers_list,
  dot.scale = 6) +
  RotatedAxis()
```

## FeaturePlot and ViolinPlot
With the cells clustered and based on the previous cell-type markers, we will select the most informative gene markers for each cluster and some well-known cell markers and project them in the UMAP. Basically, we are looking for consistency of expression of the markers across the clusters (e.g. if there are two markers for a cell type and only one of them is expressed in a cluster, we cannot reliably assign that cluster to the cell type).
```{r}
# Selected gene markers
selected_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC) %>%
  group_split()
```

```{r fig.width=7, fig.height=5}
# Feature plots
purrr::map(selected_markers, function(cluster) {
  FeaturePlot(seurat_obj,
              features = cluster$gene,
              label = TRUE,
              label.size = 2,
              ncol = 2)
})
```

## Explore gene markers

```{r fig.width=25, fig.height=6}
epidermal_genes <- c("Dcn", "Col3a1", "Col1a2")

FeaturePlot(seurat_obj,
            features = epidermal_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(seurat_obj,
            features = epidermal_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 3
)
```


```{r fig.width=25, fig.height=6}
doublets_genes <- c("Selenop", "Apoe", "Cd209d", "Cd3e", "Cd3d")

FeaturePlot(seurat_obj,
            features = doublets_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)

VlnPlot(seurat_obj,
            features = doublets_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)
```


```{r fig.width=25, fig.height=6}
cDC1_genes <- c("Irf8", "Cd24a", "Batf3", "Xcr1", "H2-Ab1")

FeaturePlot(seurat_obj,
            features = cDC1_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)

VlnPlot(seurat_obj,
            features = cDC1_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)
```

```{r fig.width=25, fig.height=6}
mast_genes <- c("Mcpt4", "Gata2", "Ms4a2", "Kit", "Fcer1a")

FeaturePlot(seurat_obj,
            features = mast_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)

VlnPlot(seurat_obj,
            features = mast_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)
```

```{r fig.width=15, fig.height=5}
cycling_genes <- c("Mki67", "Top2a", "Stmn1")

FeaturePlot(seurat_obj,
            features = cycling_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(seurat_obj,
            features = cycling_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 3 
)
```

```{r fig.width=25, fig.height=10}
neutro_genes <- c("S100a9", "S100a8", "Cd14", "Slpi", "Cebpb", "Csf3r", "Slc11a1")

FeaturePlot(seurat_obj,
            features = neutro_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)

VlnPlot(seurat_obj,
            features = neutro_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)
```

```{r fig.width=25, fig.height=10}
DC_genes <- c("H2-Aa", "H2-Eb1", "Ccr7", "Cd207", "Cd209a", "Flt3", "Cldn1", "Itgax")

FeaturePlot(seurat_obj,
            features = DC_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)

VlnPlot(seurat_obj,
            features = DC_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)
```


```{r fig.width=25, fig.height=10}
DETC_genes <- c("Nkg7", "Cd3e", "Trdc", "Tcrg-C1", "Il2rb", "Thy1", 
                "Cd4", "Cd8b1") # negative

FeaturePlot(seurat_obj,
            features = DETC_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)

VlnPlot(seurat_obj,
            features = DETC_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)
```

```{r fig.width=25, fig.height=10}
T_genes <- c("Cd3e", "Cd4", "Cd8b1", "Il17a", "Il22", "Ctla4", "Foxp3", "Gata3", "Rora", "Il7r")

FeaturePlot(seurat_obj,
            features = T_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)

VlnPlot(seurat_obj,
            features = T_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)
```


```{r fig.width=25, fig.height=10}
monomacro_genes <- c("Lyz2", "C1qb", "C1qa", "Apoe", "Selenop", "Itgam", "Plac8")

FeaturePlot(seurat_obj,
            features = monomacro_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)

VlnPlot(seurat_obj,
            features = monomacro_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)
```


# Annotation
According to the previous cell-type identities enriched in several gene markers, we will establish the following annotation:

```{r}
seurat_obj <- RenameIdents(
  seurat_obj,
 "0" = "Monocytes/Macrophages",
 "1" = "T cells",
 "2" = "Dendritic epidermal T cells",
 "3" = "DCs",
 "4" = "Neutrophils",
 "5" = "Proliferative cells",
 "6" = "Mast cells",
 "7" = "Doublets APC-T cells",
 "8" = "cDC1",
 "9" = "Non-immune cells (CD45-)"
)

# Creating new annotation variable
seurat_obj$celltypes_1 <- seurat_obj@active.ident
seurat_obj$celltypes_1 <- factor(seurat_obj$celltypes_1,
                                 levels = sort(levels(seurat_obj$celltypes_1)))
# Exploring number of celltypes
table(seurat_obj$celltypes_1)
```


## UMAP celltypes
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering_annot <- DimPlot(object = seurat_obj,
        group.by = "celltypes_1",
        label = TRUE,
        label.size = 5,
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types - level1")
gg_umap_clustering_annot

ggsave(filename = paste0(path_r_figs, "/post12_cd45pos_cells1_clustering_annotation_umap.png"),
       plot = gg_umap_clustering_annot)
```

Now, we will also explore clusters across sample.
```{r fig.width=25, fig.height=5}
DimPlot(
  object = seurat_obj,
  group.by = "celltypes_1",
  split.by = "sample_id",
  label = FALSE,
  cols = color_palette
  ) + NoLegend()
```

## Cell proportions
```{r fig.width=8, fig.height=10}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "celltypes_1",
                            group.by = "sample_id",
                            scale = "count",
                            x.reorder = c(2, 5, 3, 6, 1, 4),
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "celltypes_1",
                             group.by = "sample_id",
                             scale = "percent",
                             x.reorder = c(2, 5, 3, 6, 1, 4),
                             color.panel = color_palette)

gg_cell_num / gg_cell_prop
```

# Save Seurat object
```{r}
# Save lineage annotations
saveRDS(seurat_obj$celltypes_1,
        paste0(path_r_objects, "/post12_cd45pos_cells1_annotation.rds"))

# Save processed lognorm Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/post12_cd45pos_cells1_clustering_annotation.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/post12_cd45pos_cells1_clustering_annotation.rds"))
```


# Save Seurat objects (to subcluster)
```{r}
# Cluster 9 not using --> CD45- fraction, Epcam+ cells

# Monocytes/Macrophages
seurat_obj_monomacro <- subset(seurat_obj, 
                             idents = "Monocytes/Macrophages")
saveRDS(seurat_obj_monomacro, 
        paste0(path_r_objects, "/post12_cd45pos_monomacro.rds"))
rm(seurat_obj_monomacro)

# DCs
seurat_obj_DC <- subset(seurat_obj, 
                            idents = "DCs")
saveRDS(seurat_obj_DC,
        paste0(path_r_objects, "/post12_cd45pos_DC.rds"))
rm(seurat_obj_DC)

# T cells
seurat_obj_Tcells <- subset(seurat_obj,  
                            idents = "T cells")
saveRDS(seurat_obj_Tcells,
        paste0(path_r_objects, "/post12_cd45pos_Tcells.rds"))
rm(seurat_obj_Tcells)

# Proliferative cells
seurat_obj_prolifcells <- subset(seurat_obj,  
                            idents = "Proliferative cells")
saveRDS(seurat_obj_prolifcells,
        paste0(path_r_objects, "/post12_cd45pos_proliferativecells.rds"))
rm(seurat_obj_prolifcells)
```


# Session Info
```{r}
sessionInfo()
```
